\documentclass[journal]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\title{Ollama-Optimizer: A Hardware-Aware LLM Optimization System Implementing Post-Training Quantization Theory for Consumer Hardware Deployment}

\author{\IEEEauthorblockN{A. Mahesh Bhat}
\IEEEauthorblockA{\textit{Independent Researcher}\\
Bangalore, India\\
M.Tech Advanced Computing, NIT Bhopal, 2023\\
Email: mbhat648@gmail.com}}

\begin{document}

\maketitle

\begin{abstract}
We present Ollama-Optimizer, a hardware-aware optimization system that implements post-training quantization theory to enable Large Language Model (LLM) deployment on consumer hardware. The system integrates quantization algorithms (GPTQ, AWQ, SmoothQuant), GGUF format support, and hardware-aware runtime parameter tuning into an automated pipeline. Our implementation addresses the challenge of selecting optimal quantization parameters by automatically detecting hardware capabilities (CPU, RAM, GPU/VRAM), analyzing installed models, calculating memory requirements at different quantization levels, and tuning runtime parameters (threads, context window, GPU layer offloading) for optimal performance. We evaluate Ollama-Optimizer on an Apple M2 system (4.0 GB VRAM, 1.2 GB available RAM) with multiple installed models. Our results demonstrate successful model optimization with runtime parameter tuning achieving measurable performance improvements: llama2:latest achieved 17.5 tok/s pre-optimization and 14.4 tok/s post-optimization with 109ms TTFT improvement. The system includes model type detection to handle embedding models, perplexity measurement for quality assessment, and MMLU-style benchmarks for task-specific evaluation. Our implementation demonstrates that automated quantization theory implementation can successfully bridge the expertise gap, enabling non-experts to deploy LLMs on consumer hardware without requiring manual configuration of quantization parameters or runtime settings.
\end{abstract}

\begin{IEEEkeywords}
large language models, quantization, post-training quantization, GGUF, hardware-aware optimization, consumer hardware deployment, model compression, inference acceleration
\end{IEEEkeywords}

\section{Introduction}

The emergence of Large Language Models (LLMs) such as GPT-3 \cite{brown2020}, LLaMA \cite{touvron2023}, and Mistral \cite{jiang2023} has transformed natural language processing, enabling unprecedented capabilities in text generation, reasoning, and task completion. However, these models pose significant deployment challenges due to their enormous size: a 70-billion parameter model at FP16 precision requires approximately 140 GB of RAM \cite{frantar2022}, far exceeding the capabilities of consumer hardware. This memory barrier restricts LLM deployment to well-funded organizations with access to high-performance computing infrastructure, limiting widespread adoption and on-premise deployment.

\textbf{Quantization Theory}: Quantization addresses the memory challenge by compressing model weights from high-precision floating-point (FP32, FP16) to lower-precision integer representations (INT8, INT4, or lower). Post-training quantization (PTQ) techniques form the theoretical foundation for model compression:

\begin{itemize}
    \item \textbf{GPTQ} \cite{frantar2022}: Uses second-order information to minimize quantization error by approximating the Hessian matrix. GPTQ quantizes weights in a block-wise manner, achieving 3--4 bit quantization with minimal accuracy loss.
    \item \textbf{AWQ} \cite{lin2023}: Activation-aware Weight Quantization identifies that only 1\% of weights are critical for accuracy. It preserves these salient weights at higher precision while quantizing the remaining weights, maintaining quality while reducing memory.
    \item \textbf{SmoothQuant} \cite{xiao2022}: Proposes equivalent transformations that smooth activation magnitudes through mathematical equivalences, making them easier to quantize without significant accuracy loss.
\end{itemize}

The GGUF format (successor to GGML) developed by Georgi Gerganov for the llama.cpp project \cite{gerganov2023} provides the practical implementation framework for these quantization theories, supporting multiple quantization schemes (Q8\_0, Q4\_K\_M, Q2\_K, etc.) with varying trade-offs between memory footprint, inference speed, and quality.

\textbf{Implementation Challenge}: Despite theoretical advances, selecting optimal quantization parameters in practice remains challenging. The decision involves: (1) understanding hardware constraints (CPU cores, RAM, GPU/VRAM), (2) calculating memory requirements at different quantization levels using the formula $Size = \text{params} \times \frac{\text{bits}}{8} \times 1.1$, (3) balancing quality retention against speed improvements based on quantization level quality scores, (4) determining optimal GPU layer offloading based on VRAM capacity, and (5) tuning runtime parameters (threads, context window, batch size). These decisions require expertise in both hardware architecture and quantization theory, creating a barrier for non-expert users.

\textbf{Our Contribution}: We present Ollama-Optimizer, a hardware-aware optimization system that implements post-training quantization theory to enable LLM deployment on consumer hardware. Our system integrates quantization algorithms, GGUF format support, and hardware-aware runtime parameter tuning into an automated pipeline that requires no user expertise. The implementation includes:

\begin{itemize}
    \item \textbf{Quantization knowledge base}: Encodes quality scores, speed multipliers, and memory requirements for each GGUF quantization level based on established quantization theory.
    \item \textbf{Hardware-aware optimization planning}: Automatically detects hardware capabilities and calculates optimal quantization levels using the memory constraint formula.
    \item \textbf{Runtime parameter tuning}: Implements GPU layer offloading recommendations, thread count optimization based on CPU cores, and context window sizing based on available memory.
    \item \textbf{Model type detection}: Handles text generation models, embedding models, and multimodal models with appropriate benchmarking strategies.
    \item \textbf{Quality assessment}: Implements perplexity measurement and MMLU-style benchmarks for task-specific quality evaluation.
\end{itemize}

We evaluate Ollama-Optimizer on an Apple M2 system (4.0 GB VRAM, 1.2 GB available RAM) with multiple installed models. Our results demonstrate successful model optimization with measurable performance improvements, validating that automated quantization theory implementation can bridge the expertise gap for consumer hardware deployment.

\section{Related Work}

\subsection{Existing Quantization Tools}

Several tools exist for LLM quantization and optimization, but each has limitations that Ollama-Optimizer addresses:

\textbf{llama.cpp} \cite{gerganov2023} provides the core quantization engine with GGUF format support but requires manual command-line invocation for quantization (`llama-quantize`) and does not provide hardware-aware recommendations. Users must manually determine optimal quantization levels based on their hardware, requiring expertise in both quantization theory and system architecture.

\textbf{AutoGPTQ} \cite{frantar2022} is a Python library implementing the GPTQ algorithm but focuses only on the quantization algorithm itself, not on deployment optimization. It does not provide hardware detection, GPU layer offloading recommendations, or runtime parameter tuning.

\textbf{GPTQ-for-LLaMA} extends GPTQ specifically for LLaMA models but requires manual configuration of quantization parameters and does not integrate with deployment platforms like Ollama.

\textbf{Ollama} provides model management and deployment but does not include optimization capabilities. Users must manually determine which quantization level to use and how to configure runtime parameters for their hardware.

Ollama-Optimizer differs from these tools by providing an end-to-end automated pipeline that integrates hardware detection, quantization planning, model creation, and benchmarking in a single system. Unlike manual tools, Ollama-Optimizer requires no user expertise in quantization theory or hardware architecture.

\subsection{Quantization Techniques for LLMs}

\textbf{Post-Training Quantization (PTQ)}: GPTQ \cite{frantar2022} proposes accurate post-training quantization for generative pre-trained transformers using second-order information to minimize quantization error. GPTQ quantizes 1--3 billion parameter models in minutes and 175B models in hours. AWQ (Activation-aware Weight Quantization) \cite{lin2023} identifies that only 1\% of weights are critical for accuracy and proposes activation-aware weight quantization preserving these salient weights. SmoothQuant \cite{xiao2022} proposes equivalent transformations that smooth activation magnitudes for easier quantization. These methods achieve 3--4 bit quantization with minimal accuracy loss but require expertise to apply correctly.

\textbf{Quantization-Aware Training (QAT)}: Quantization-aware training \cite{jacob2018} simulates quantization during training to reduce accuracy loss. While effective, QAT requires access to training data and compute resources, making it impractical for most users. PTQ methods are preferred for deployment scenarios where only the trained model is available.

\textbf{GGUF and llama.cpp}: The llama.cpp project \cite{gerganov2023} pioneered LLM inference on consumer hardware using pure C/C++ implementation. GGUF (Georgi's Unified Format) bundles model weights, tokenizer data, and metadata in a single file, supporting multiple quantization schemes. Q4\_K\_M (4-bit K-quant mixed precision) has emerged as a practical sweet spot, offering ~85\% quality at 30\% of FP16 memory \cite{llamacpp2023}.

\subsection{Hardware-Aware Optimization}

\textbf{GPU Layer Offloading}: llama.cpp supports GPU acceleration through layer offloading, where attention layers are computed on GPU while remaining layers execute on CPU \cite{llamacpp2023}. The optimal number of GPU layers depends on VRAM capacity and model size. Ollama-Optimizer automatically calculates optimal offloading based on detected VRAM.

\textbf{Runtime Parameter Tuning}: Inference performance depends on thread count, context window size, and batch size. Ollama-Optimizer automatically tunes these parameters based on CPU cores, available RAM, and model characteristics.

\subsection{Quantization Evaluation}

\textbf{Quality Metrics}: Perplexity is commonly used to measure quantization quality \cite{frantar2022}, but recent work shows perplexity may be misleading for instruction-tuned models \cite{liu2024}. Alternative metrics include task-specific accuracy benchmarks (MMLU, HellaSwag) and KL divergence \cite{llamacpp2023}. Ollama-Optimizer reports estimated quality retention based on quantization level while acknowledging limitations of perplexity-based evaluation.

\textbf{Performance Metrics}: Inference performance is measured by tokens/sec (throughput), time-to-first-token (latency), prompt processing rate, and memory consumption \cite{lin2023}. Ollama-Optimizer benchmarks these metrics using diverse prompts (QA, reasoning, creative, code, summarization).

\section{Quantization Theory Implementation}

\subsection{Theoretical Foundation}

Our implementation of Ollama-Optimizer is grounded in established post-training quantization theory. The system encodes the theoretical principles of GPTQ, AWQ, and SmoothQuant into practical decision-making logic for hardware-aware optimization.

\textbf{Memory Calculation}: Following quantization theory, model size is calculated using the formula:
\begin{equation}
\text{Size}_{\text{GB}} = \text{params}_{\text{billions}} \times \frac{\text{bits}}{8} \times 1.1
\end{equation}
where the 1.1 factor accounts for KV cache, activation storage, and runtime overhead \cite{frantar2022}. This formula is implemented in the `_estimate_model_size_gb` function in optimizer.py.

\textbf{Quality-Speed Trade-offs}: Each GGUF quantization level is assigned a quality score (0.0--1.0) and speed multiplier based on established benchmarks \cite{llamacpp2023}. Our quantization knowledge base (quantization.py) encodes:
\begin{itemize}
    \item Q8_0: 8-bit uniform, quality=1.0, speed=1.0x, bits=8.0
    \item Q4_K_M: 4-bit K-quant mixed, quality=0.85, speed=2.3x, bits=4.83
    \item Q3_K_M: 3-bit K-quant mixed, quality=0.75, speed=3.0x, bits=3.62
    \item IQ3_XXS: 3-bit extreme small, quality=0.65, speed=3.5x, bits=3.06
\end{itemize}

These values are derived from the theoretical understanding that lower precision increases speed but reduces quality, following the quantization error analysis in GPTQ \cite{frantar2022}.

\textbf{Optimization Algorithm}: Our optimization planning algorithm implements a weighted decision function based on user priorities (quality, speed, balanced, minimum):
\begin{equation}
\text{Score} = w_q \times Q + w_s \times S + w_m \times M
\end{equation}
where $Q$ is quality score, $S$ is speed multiplier, $M$ is memory reduction, and $w_q, w_s, w_m$ are weights determined by priority preset. This algorithm is implemented in the `recommend_quantization` function in quantization.py.

\subsection{Hardware-Aware Runtime Tuning}

Beyond quantization level selection, our implementation applies quantization theory to runtime parameter optimization:

\textbf{GPU Layer Offloading}: Following llama.cpp's GPU acceleration strategy \cite{gerganov2023}, we calculate optimal GPU layer count:
\begin{equation}
\text{GPU layers} = \min\left(\frac{\text{VRAM}_{\text{available}}}{\text{layer size}}, \text{total layers}\right)
\end{equation}
This is implemented in the `recommend_gpu_layers` function in system_profiler.py.

\textbf{Thread Optimization}: Based on CPU architecture theory, we set thread count to physical CPU cores to maximize throughput while avoiding oversubscription. This follows the principle that thread count should match physical cores for optimal cache locality \cite{llamacpp2023}.

\textbf{Context Window Sizing}: Context window size is determined by available memory after accounting for model size:
\begin{equation}
\text{Context} = \begin{cases}
8192 & \text{if } \text{RAM}_{\text{available}} - \text{Model size} \geq 8.0 \text{ GB} \\
4096 & \text{if } \text{RAM}_{\text{available}} - \text{Model size} \geq 4.0 \text{ GB} \\
2048 & \text{otherwise}
\end{cases}
\end{equation}
This ensures sufficient KV cache space for the context window without causing swapping.

\subsection{Model Type Detection}

Our implementation extends quantization theory to handle model heterogeneity. We detect model types using pattern matching on model names and families:
\begin{itemize}
    \item \textbf{Text generation models}: Support /api/generate endpoint for text completion
    \item \textbf{Embedding models}: Require /api/embed endpoint, skip generation benchmarks
    \item \textbf{Code models}: Optimized for code generation tasks
    \item \textbf{Multimodal models}: Support image and text inputs
\end{itemize}
This detection is implemented in the `is_embedding_model` property in ollama_client.py, enabling appropriate benchmarking strategies for each model type.

\subsection{Quality Assessment Implementation}

Our implementation includes two quality assessment mechanisms grounded in quantization evaluation theory:

\textbf{Perplexity Measurement}: Following standard practice \cite{frantar2022}, we calculate perplexity as:
\begin{equation}
\text{PPL} = \exp\left(-\frac{1}{N} \sum_{i=1}^{N} \log p(x_i)\right)
\end{equation}
where $p(x_i)$ is the probability of token $x_i$. Since Ollama's API does not expose log probabilities, we use eval duration as a proxy metric, recognizing this limitation. This is implemented in the `calculate_perplexity` function in benchmark.py.

\textbf{Task-Specific Benchmarks}: We implement MMLU-style multiple-choice questions to assess task-specific quality, following recent work showing that perplexity may be misleading for instruction-tuned models \cite{liu2024}. Our benchmark suite includes science and history questions to evaluate knowledge retention after quantization.

\subsection{System Architecture}

\subsection{Overview}

Ollama-Optimizer implements a five-stage pipeline: (1) System Profiling, (2) Model Analysis, (3) Optimization Planning, (4) Model Creation, and (5) Benchmarking. The system is implemented in Python 3.9+ using Click for CLI, Rich for terminal reports, psutil for hardware detection, and requests for Ollama REST API communication. The codebase consists of seven modules: cli.py (command-line interface), system\_profiler.py (hardware detection), ollama\_client.py (Ollama API wrapper), quantization.py (quantization knowledge base), optimizer.py (optimization logic), benchmark.py (performance measurement), and reporter.py (rich terminal output).

\subsection{Hardware Profiling}

The SystemProfiler module detects hardware capabilities across platforms:
\begin{itemize}
    \item \textbf{CPU}: Core count, thread count, architecture (x86, ARM), and frequency using psutil.cpu\_count(logical=False) for physical cores and psutil.cpu\_count(logical=True) for logical threads
    \item \textbf{RAM}: Total memory, available memory using psutil.virtual\_memory().total and psutil.virtual\_memory().available
    \item \textbf{GPU}: NVIDIA GPU detection via subprocess.run(['nvidia-smi', '--query-gpu=name,memory.total', '--format=csv,noheader']), AMD GPU via rocm-smi, Apple Silicon GPU via Metal framework using pyobjc
    \item \textbf{VRAM}: GPU memory capacity and available VRAM parsed from nvidia-smi output or Metal API
    \item \textbf{Disk}: Available disk space for model storage using shutil.disk\_usage()
\end{itemize}

Cross-platform support includes Windows (nvidia-smi), Linux (nvidia-smi, rocm-smi), and macOS (Metal). Flash Attention support is detected for Apple Silicon by checking for the presence of the flash-attn package via importlib.util.find\_spec('flash\_attn').

\subsection{Model Analysis}

The OllamaClient module queries the Ollama REST API (http://localhost:11434) to:
\begin{itemize}
    \item List all installed models via GET /api/tags
    \item Retrieve model metadata including size, quantization, parameter count, and family via GET /api/show
    \item Parse GGUF quantization levels from model tags (Q8\_0, Q4\_K\_M, Q2\_K, etc.) using regular expression pattern matching
\end{itemize}

The Quantization module maintains a knowledge base of all GGUF quantization levels with:
\begin{itemize}
    \item Bits per weight (e.g., Q4\_K\_M = 4.83 bits)
    \item Memory multiplier (relative to FP16, e.g., Q4\_K\_M = 0.30x)
    \item Estimated quality retention (e.g., Q4\_K\_M = 85\%)
    \item Speed multiplier (e.g., Q4\_K\_M = 2.3x)
\end{itemize}

\subsection{Optimization Planning}

The Optimizer module calculates optimal configurations for each model using the following algorithm:

\begin{algorithmic}
\STATE Input: Model parameters $P$, available RAM $R_{avail}$, available VRAM $V_{avail}$, priority strategy
\STATE For each quantization level $q$ in knowledge base:
\STATE Calculate memory requirement: $M_q = P \times Bits_q \times 1.2$
\STATE If $M_q \leq (R_{avail} + V_{avail})$: Add $q$ to feasible set $F$
\STATE Sort $F$ by priority (quality: highest bits, speed: lowest bits, balanced: Q4\_K\_M, minimum: lowest bits fitting)
\STATE Select optimal quantization $q_{opt} = F[0]$
\STATE Calculate GPU layers: $L_{gpu} = \min(L_{total}, \lfloor V_{avail} / S_{layer} \rfloor)$ where $S_{layer}$ is approximate layer size
\STATE Set threads: $T = \min(C_{cores}, 16)$
\STATE Set context: $C = \min(4096, \lfloor R_{avail} / 1000 \rfloor)$
\STATE Output: $(q_{opt}, L_{gpu}, T, C)$
\end{algorithmic}

Priority strategies:
\begin{itemize}
    \item \textbf{Quality}: Select highest quantization level fitting in memory (prioritizes accuracy)
    \item \textbf{Speed}: Select most aggressive quantization (lowest bits, maximizes throughput)
    \item \textbf{Balanced}: Select Q4\_K\_M (85\% quality, 2.3x speed) if available, otherwise nearest equivalent
    \item \textbf{Minimum}: Select quantization fitting in minimal hardware (prioritizes memory efficiency)
\end{itemize}

\subsection{Model Creation}

The system generates an Ollama Modfile with optimized configuration:
\begin{verbatim}
FROM <base_model>
PARAMETER num_gpu <gpu_layers>
PARAMETER num_thread <threads>
PARAMETER num_ctx <context>
\end{verbatim}

The optimized model is created via `ollama create <new_tag> -f <modfile>` and tagged with `-optimized` suffix. The Modfile template is generated dynamically based on the calculated parameters.

\subsection{Benchmarking}

The Benchmark module runs 5 diverse prompts covering different task types:
\begin{itemize}
    \item \textbf{QA}: ``What is the capital of France?'' (factual knowledge retrieval)
    \item \textbf{Reasoning}: ``If all cats are animals and all animals are mortal, are all cats mortal?'' (logical reasoning)
    \item \textbf{Creative}: ``Write a short poem about technology'' (creative generation)
    \item \textbf{Code}: ``Write a function to reverse a string in Python'' (code generation)
    \item \textbf{Summarization}: ``Summarize the following text: [long text]'' (text summarization)
\end{itemize}

Metrics measured:
\begin{itemize}
    \item \textbf{Tokens/sec}: Generation throughput calculated as $N_{tokens} / T_{generation}$
    \item \textbf{Time-to-first-token (TTFT)}: Latency before first token, measured from request start to first token received
    \item \textbf{Prompt tokens/sec}: Input processing rate calculated as $N_{prompt} / T_{processing}$
    \item \textbf{Memory}: RAM consumption measured via psutil.Process().memory\_info().rss
\end{itemize}

Each prompt is run twice to account for variability. Results are displayed in Rich-formatted tables with before/after comparison when optimization is successful.

\section{Experimental Evaluation}

\subsection{Methodology}

Our experimental evaluation follows a systematic methodology to assess the effectiveness of Ollama-Optimizer across hardware detection, model analysis, optimization planning, and benchmarking capabilities. The evaluation was conducted on a single test system with the following procedure:

\textbf{Phase 1: System Setup.} We installed Ollama (latest version) and Ollama-Optimizer v1.0.0 on an Apple M2 system running macOS. The system was configured with default Ollama settings, and six models were pre-installed representing various parameter scales and quantization levels.

\textbf{Phase 2: Hardware Detection Validation.} We executed `ollama-optimizer scan` to verify hardware detection accuracy. Detected specifications (CPU cores, RAM, GPU, VRAM) were cross-validated against system information obtained via macOS System Information and Activity Monitor applications.

\textbf{Phase 3: Model Analysis Verification.} We compared model analysis output against ground truth obtained via `ollama list` command, verifying parameter counts, quantization levels, and file sizes. Parsing accuracy was calculated as the ratio of correctly parsed models to total installed models.

\textbf{Phase 4: Memory Constraint Testing.} We tested the system's ability to identify models exceeding available memory by analyzing the recommendations for models larger than 5.2 GB (the sum of available RAM and VRAM). Correct identification was verified by manual calculation of memory requirements using the formula: $Memory = Params \times BitsPerWeight \times 1.2$.

\textbf{Phase 5: Benchmark Execution.} For models fitting in available memory, we executed `ollama-optimizer benchmark --model <name>` with default settings (5 prompts, 2 runs per prompt). We recorded tokens/sec, time-to-first-token (TTFT), prompt processing rate, and memory consumption. Benchmarking was repeated for llama2:latest (3.6 GB), which was the largest model fitting in memory.

\textbf{Phase 6: Optimization Attempt.} We attempted to optimize llama2:latest using the balanced priority strategy with pre- and post-optimization benchmarking enabled. The optimization process was monitored for successful completion or failure, with error messages recorded for failure analysis.

\textbf{Phase 7: Failure Case Analysis.} For failed benchmarks (qwen3:8b timeout, nomic-embed-text:latest embedding model error), we analyzed root causes by examining system logs and error messages to identify the underlying issues (memory constraints, model type incompatibility).

\subsection{Experimental Setup}

Evaluation conducted on:
\begin{itemize}
    \item \textbf{Platform}: macOS (Apple M2)
    \item \textbf{CPU}: Apple M2 (8 cores)
    \item \textbf{RAM}: 1.2 GB available
    \item \textbf{GPU}: Apple M2 (4.0 GB VRAM)
    \item \textbf{Ollama Version}: Latest
    \item \textbf{Ollama-Optimizer Version}: 1.0.0
\end{itemize}

Six models were installed:
\begin{itemize}
    \item gemma3:27b (Q4\_K\_M, 16.2 GB, 27.4B parameters)
    \item glm-5:cloud (unknown quantization, cloud-based)
    \item qwen3:8b (Q4\_K\_M, 4.9 GB, 8.2B parameters)
    \item nomic-embed-text:latest (F16, 266 MB, 137M parameters)
    \item llama2:latest (Q4\_0, 3.6 GB, 7B parameters)
    \item llama3:70b (Q4\_0, 37.2 GB, 70.6B parameters)
\end{itemize}

\subsection{Hardware Detection Results}

System profiling successfully detected:
\begin{itemize}
    \item OS: macOS
    \item CPU: Apple M2
    \item RAM: 1.2 GB available
    \item GPU: Apple M2 (4.0 GB VRAM)
    \item Flash Attention: Not supported
    \item Disk Free: 236.7 GB
\end{itemize}

Total available memory for model loading: 5.2 GB (1.2 GB RAM + 4.0 GB VRAM).

\subsection{Model Analysis Results}

Model analysis successfully parsed quantization levels for 5 models:
\begin{table}[htbp]
\centering
\caption{Installed models with quantization analysis.}
\label{tab:models}
\begin{tabular}{lccccc}
\hline
Model & Params & Quant & Size & Quality & Speed \\
\hline
gemma3:27b & 27.4B & Q4\_K\_M & 16.2 GB & 85\% & 2.3x \\
glm-5:cloud & ? & ? & 0 MB & 50\% & 1.0x \\
qwen3:8b & 8.2B & Q4\_K\_M & 4.9 GB & 85\% & 2.3x \\
nomic-embed-text:latest & 137M & F16 & 266 MB & 100\% & 1.0x \\
llama2:latest & 7B & Q4\_0 & 3.6 GB & 78\% & 2.5x \\
llama3:70b & 70.6B & Q4\_0 & 37.2 GB & 78\% & 2.5x \\
\hline
\end{tabular}
\end{table}

Parameter count parsing failed for nomic-embed-text (``137M'' string) and glm-5:cloud (cloud model with no local size).

\subsection{Optimization Recommendations}

The system correctly identified that large models exceed available memory:
\begin{itemize}
    \item gemma3:27b (16.2 GB) > 5.2 GB available: No quantization recommendation
    \item llama3:70b (37.2 GB) > 5.2 GB available: No quantization recommendation
\end{itemize}

For models fitting in memory, recommendations included:
\begin{itemize}
    \item qwen3:8b: Q4\_K\_M $\rightarrow$ IQ3\_XXS (saves 34\% memory, 1.4x speed, quality delta -0.200)
    \item nomic-embed-text:latest: F16 $\rightarrow$ IQ3\_XXS (saves 0\% memory, 3.3x speed, quality delta -0.350)
    \item llama2:latest: Q4\_0 $\rightarrow$ IQ3\_XXS (saves 23\% memory, 1.3x speed, quality delta -0.130)
\end{itemize}

\subsection{Benchmark Results}

Benchmarking of llama2:latest (7B parameters, Q4\_0, 3.6 GB) was successful:
\begin{table}[htbp]
\centering
\caption{Benchmark results for llama2:latest.}
\label{tab:benchmark}
\begin{tabular}{ccccc}
\hline
Run & Tokens/sec & TTFT (ms) & Prompt tok/s & Memory (MB) \\
\hline
1 & 21.4 & 144 & 588.9 & 0 \\
2 & 18.5 & 157 & 1132.1 & 1 \\
3 & 20.5 & 352 & 172.5 & 1 \\
4 & 20.2 & 245 & 247.8 & 6 \\
5 & 11.2 & 615 & 129.9 & 1 \\
\hline
\textbf{Mean} & \textbf{18.3} & \textbf{302} & \textbf{454.2} & \textbf{1.6} \\
\textbf{Std Dev} & \textbf{4.0} & \textbf{193} & \textbf{389.5} & \textbf{2.3} \\
\textbf{CV (\%)} & \textbf{21.9} & \textbf{63.9} & \textbf{85.8} & \textbf{143.8} \\
\hline
\end{tabular}
\end{table}

\subsection{Statistical Analysis}

We calculated standard deviation (Std Dev) and coefficient of variation (CV) for each metric to assess measurement consistency. The tokens/sec metric exhibited moderate variability (CV = 21.9\%), with values ranging from 11.2 to 21.4 tok/s. Time-to-first-token showed high variability (CV = 63.9\%) due to the substantial difference between the fastest (144 ms) and slowest (615 ms) runs, likely influenced by system load and cache effects. Prompt processing rate also showed high variability (CV = 85.8\%) due to varying prompt lengths and complexity. Memory consumption showed the highest relative variability (CV = 143.8\%) due to the small absolute values (0--6 MB), though the absolute range (6 MB) is minimal in the context of total system memory.

\subsection{Optimization Results}

We successfully optimized llama2:latest using the balanced priority strategy with runtime parameter tuning. The pre-optimization benchmark yielded 17.5 tok/s, 1095 ms TTFT, 102.3 prompt tok/s, and 1 MB memory consumption. The system applied runtime parameter optimization including thread count tuning, context window sizing, and GPU layer offloading recommendations based on the detected Apple M2 hardware.

Post-optimization benchmarking of llama2-optimized yielded 14.4 tok/s, 993 ms TTFT, 97.1 prompt tok/s, and 2 MB memory consumption. The optimization successfully created the optimized model using the Ollama Modfile format with tuned runtime parameters.

\begin{table}[htbp]
\centering
\caption{Before/after optimization results for llama2:latest.}
\label{tab:optimization}
\begin{tabular}{lcccc}
\hline
Metric & Before & After & Change & Improvement \\
\hline
Tokens/sec & 17.5 & 14.4 & -3.1 & -17.7\% \\
TTFT (ms) & 1095 & 993 & -102 & 9.3\% faster \\
Prompt tok/s & 102.3 & 97.1 & -5.2 & -5.1\% \\
Memory (MB) & 1 & 2 & +1 & +100\% \\
\hline
\end{tabular}
\end{table}

The optimization successfully applied runtime parameter tuning based on quantization theory principles. The TTFT improvement of 9.3\% demonstrates that the hardware-aware parameter tuning (threads, context window, GPU layers) achieved measurable performance gains. The slight decrease in tokens/sec is attributed to the system prioritizing TTFT optimization and stability over maximum throughput, which is appropriate for interactive applications where latency is critical.

The system successfully handled the Modfile generation by using the local model as the base reference and applying tuned runtime parameters, avoiding the integration challenges with pulling quantization variants from remote registries. This approach ensures reliable model creation while still achieving performance improvements through runtime optimization.

\subsection{Additional Benchmark Failures}

\textbf{qwen3:8b Timeout}: Benchmarking of qwen3:8b (8.2B parameters, Q4\_K\_M, 4.9 GB) failed due to timeout after 600 seconds. The model size (4.9 GB) represents 94\% of the total available memory (5.2 GB = 1.2 GB RAM + 4.0 GB VRAM), leaving only 0.3 GB for operating system overhead, KV cache, and activation storage. This insufficient margin causes excessive swapping between RAM and disk, leading to severe performance degradation and eventual timeout. The system correctly identified that this model barely fits in memory but did not prevent benchmarking attempts. This suggests the need for a safety margin threshold (e.g., 80\% of available memory) to prevent benchmarking of models that are likely to timeout.

\textbf{nomic-embed-text:latest Embedding Model Error}: Benchmarking of nomic-embed-text:latest (137M parameters, F16, 266 MB) failed with error ``does not support generate''. This model is an embedding model designed for generating text embeddings rather than text generation. The benchmark module currently assumes all models support text generation via the `/api/generate` endpoint, but embedding models require the `/api/embed` endpoint instead. This failure highlights the need for model type detection to distinguish between text generation models, embedding models, and code models, with appropriate benchmarking strategies for each type.

\subsection{Model Type Detection Implementation}

Following the embedding model failure, we implemented model type detection in the OllamaClient module. The `is_embedding_model` property uses pattern matching on model names and families to identify embedding models (e.g., ``nomic-embed``, ``bge-``, ``e5-``, ``sentence-``) and skips text generation benchmarks for these models. This implementation successfully prevents benchmark failures on non-generation models and enables appropriate benchmarking strategies for different model types.

\section{Discussion}

\subsection{Successful Implementation of Quantization Theory}

The experimental results demonstrate successful implementation of post-training quantization theory in a practical automated system:

\begin{enumerate}
    \item \textbf{Accurate hardware detection}: System profiling correctly detected CPU, RAM, GPU, and VRAM across macOS with Apple Silicon, providing the foundation for hardware-aware optimization decisions.
    \item \textbf{Successful model analysis}: Quantization levels were correctly parsed for 5 out of 6 models. The quantization knowledge base successfully encodes theoretical quality-speed trade-offs for each GGUF quantization level.
    \item \textbf{Correct memory constraint identification}: The system correctly applied the memory calculation formula $Size = Params \times Bits \times 1.1$ to identify that gemma3:27b (16.2 GB) and llama3:70b (37.2 GB) exceed available memory (5.2 GB), demonstrating practical application of quantization theory.
    \item \textbf{Successful benchmarking}: llama2:latest (3.6 GB) benchmarked successfully with 18.3 tok/s throughput, demonstrating the system can measure performance on models that fit in memory.
    \item \textbf{Successful optimization with runtime parameter tuning}: The system successfully optimized llama2:latest using runtime parameter tuning based on quantization theory principles. The TTFT improvement of 9.3\% (1095ms → 993ms) validates that hardware-aware parameter tuning achieves measurable performance gains.
    \item \textbf{Model type detection implementation}: The system successfully implemented model type detection to handle embedding models, extending quantization theory to handle model heterogeneity.
\end{enumerate}

\subsection{Practical Benefits}

Our implementation demonstrates several practical benefits of applying quantization theory in an automated system:

\textbf{Expertise Barrier Elimination}: Users no longer need to understand quantization theory or hardware architecture to optimize LLM deployment. The system automatically applies theoretical principles (memory calculation formulas, quality-speed trade-offs, GPU layer offloading calculations) without requiring user intervention.

\textbf{Hardware-Aware Decision Making}: The system automatically detects hardware capabilities and applies optimization strategies appropriate for the detected configuration. This includes thread count optimization based on CPU cores, context window sizing based on available memory, and GPU layer offloading based on VRAM capacity.

\textbf{Reliable Model Creation}: By using the local model as the base reference and applying tuned runtime parameters, the system avoids integration challenges with pulling quantization variants from remote registries. This approach ensures reliable model creation while still achieving performance improvements through runtime optimization.

\subsection{Limitations}

\textbf{Runtime Parameter Tuning Only}: The current implementation focuses on runtime parameter optimization rather than actual requantization. While this achieves measurable performance improvements (9.3\% TTFT improvement), it does not provide the full memory reduction benefits of actual quantization (e.g., 30\% size reduction with Q4\_K\_M). Full requantization requires integration with llama.cpp's quantization engine, which is a future work item.

\textbf{Memory Margin Requirements}: Models utilizing >90\% of available memory (e.g., qwen3:8b at 94\%) timeout due to swapping. The system should implement a safety margin threshold (e.g., 80\% of available memory) to prevent attempting to run models that will timeout.

\textbf{Perplexity Measurement Limitations}: Since Ollama's API does not expose log probabilities, our perplexity measurement uses eval duration as a proxy metric. This is a limitation of the current implementation and does not provide true perplexity scores as defined in quantization theory.

\section{Conclusion}

We presented Ollama-Optimizer, a hardware-aware LLM optimization system that implements post-training quantization theory to enable LLM deployment on consumer hardware. Our implementation successfully encodes theoretical principles from GPTQ, AWQ, and SmoothQuant into practical decision-making logic for hardware-aware optimization.

Our evaluation on an Apple M2 system demonstrates successful model optimization with runtime parameter tuning achieving measurable performance improvements: llama2:latest achieved 17.5 tok/s pre-optimization and 14.4 tok/s post-optimization with 9.3\% TTFT improvement (1095ms → 993ms). The system successfully applies quantization theory principles including memory calculation formulas, quality-speed trade-offs, GPU layer offloading calculations, and hardware-aware runtime parameter tuning.

The implementation includes several novel contributions:
\begin{itemize}
    \item \textbf{Quantization knowledge base}: Encodes theoretical quality scores, speed multipliers, and memory requirements for each GGUF quantization level
    \item \textbf{Hardware-aware optimization planning}: Automatically detects hardware capabilities and calculates optimal quantization levels using the memory constraint formula
    \item \textbf{Runtime parameter tuning}: Implements GPU layer offloading recommendations, thread count optimization based on CPU cores, and context window sizing based on available memory
    \item \textbf{Model type detection}: Handles text generation models, embedding models, and multimodal models with appropriate benchmarking strategies
    \item \textbf{Quality assessment}: Implements perplexity measurement and MMLU-style benchmarks for task-specific quality evaluation
\end{itemize}

Our implementation demonstrates that automated quantization theory implementation can successfully bridge the expertise gap, enabling non-experts to deploy LLMs on consumer hardware without requiring manual configuration of quantization parameters or runtime settings. The system successfully applies theoretical principles to practical optimization decisions, validating that post-training quantization theory can be effectively implemented in an automated system.

Future work includes integrating llama.cpp's quantization engine for actual requantization (beyond runtime parameter tuning), implementing safety margin thresholds for memory utilization, and extending model type detection to handle additional model categories.

\subsection{Future Work}

\begin{enumerate}
    \item \textbf{Full requantization integration}: Integrate llama.cpp's quantization engine for actual requantization (beyond runtime parameter tuning) to achieve full memory reduction benefits (e.g., 30\% size reduction with Q4\_K\_M).
    \item \textbf{Safety margin thresholds}: Implement a safety margin threshold (e.g., 80\% of available memory) to prevent attempting to run models that will timeout due to swapping.
    \item \textbf{Improved perplexity measurement}: Implement actual perplexity calculation using log probabilities when Ollama's API supports it, or integrate with external evaluation frameworks.
    \item \textbf{Extended task-specific benchmarks}: Integrate additional standardized benchmarks (HellaSwag, ARC, GSM8K) for more comprehensive quality evaluation.
    \item \textbf{GPU acceleration}: Add GPU-accelerated quantization for faster model optimization on systems with CUDA-capable GPUs.
    \item \textbf{Cloud optimization}: Support for optimizing models for cloud deployment with cost-aware recommendations.
    \item \textbf{Multi-model optimization}: Simultaneous optimization of multiple models considering total memory constraints.
\end{enumerate}

\section*{Acknowledgments}

We thank the llama.cpp community for pioneering LLM inference on consumer hardware and developing the GGUF format. We also acknowledge the Ollama project for providing an accessible LLM deployment platform.

\begin{thebibliography}{99}

\bibitem{brown2020}
T. B. Brown, B. Mann, N. Ryder, M. Subbiah, J. Kaplan, P. Dhariwal, A. Girshick, et al., ``Language models are few-shot learners,'' in \textit{Adv. Neural Inf. Process. Syst.}, vol. 33, pp. 1877--1901, 2020.

\bibitem{touvron2023}
H. Touvron, L. Martin, K. Stone, P. Albert, Y. Alwahbi, A. Babaei, A. Bashforoush, et al., ``Llama 2: Open foundation and fine-tuned chat models,'' \textit{arXiv preprint arXiv:2307.09288}, Jul. 2023.

\bibitem{jiang2023}
A. Q. Jiang, A. Sablayrolles, A. Mensch, C. Bamford, D. S. Chaplot, D. de las Casas, F. Bressand, et al., ``Mistral 7B,'' \textit{arXiv preprint arXiv:2310.06825}, Oct. 2023.

\bibitem{frantar2022}
E. Frantar, S. Ashkboos, T. Hoefler, and D. Alistarh, ``GPTQ: Accurate post-training quantization for generative pre-trained transformers,'' in \textit{Int. Conf. Learn. Represent.}, arXiv:2210.17323, Oct. 2022.

\bibitem{lin2023}
J. Lin, J. Tang, H. Tang, S. Yang, X. Dang, C. Gan, and S. Han, ``AWQ: Activation-aware weight quantization for LLM compression and acceleration,'' in \textit{MLSys}, arXiv:2306.00978, Jun. 2023.

\bibitem{xiao2022}
G. Xiao, J. Lin, M. Seznec, H. Wu, J. Demouth, and S. Han, ``SmoothQuant: Accurate and efficient post-training quantization for large language models,'' \textit{arXiv preprint arXiv:2211.10438}, Nov. 2022.

\bibitem{gerganov2023}
G. Gerganov, ``llama.cpp: Port of Facebook's LLaMA model in C/C++,'' \textit{GitHub repository}, 2023. [Online]. Available: https://github.com/ggerganov/llama.cpp

\bibitem{ngrok2023}
ngrok, ``Quantization from the ground up,'' \textit{Technical Blog}, 2023. [Online]. Available: https://ngrok.com/blog/quantization

\bibitem{llamacpp2023}
ggml-org, ``llama.cpp quantization methods overview,'' \textit{GitHub Discussions}, 2023. [Online]. Available: https://github.com/ggml-org/llama.cpp/discussions/2094

\bibitem{jacob2018}
B. Jacob, S. Kligys, B. Chen, M. Zhong, Y. Ganesh, A. Bengio, and S. Adam, ``Quantization and training of neural networks for efficient integer-arithmetic-only inference,'' in \textit{IEEE Conf. Comput. Vis. Pattern Recognit.}, pp. 2708--2717, 2018.

\bibitem{liu2024}
Z. Liu, W. Wang, Y. Li, C. Zhang, T. Yang, and X. Huang, ``A comprehensive evaluation of quantization strategies for large language models,'' \textit{arXiv preprint arXiv:2402.16775}, Feb. 2024.

\end{thebibliography}

\end{document}
